---
title: "ggplot2_walkthrough_palmer_penguins"
author: "J. Mular"
date: "2023-04-22"
output: html_document
description: ggplot2 package visualizations, geom_jitter, geom_point, geom_smooth
---
Revisions:
4/22/23: Tutorial split from master file into separate tutorials for ease


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Setting Up Environment
If you have not already set up your environment by installing and loading your ggplot2 packages please do so now, if you need a refresher please see the R_basics_walkthrough_palmer_penguins.rmd file

```{r}
library(tidyverse)
library(palmerpenguins)
penguins<- penguins %>% filter(!is.na(sex))
```


#Making Plots Using ggplot
Using our example of counting species that are on the island

Let's try and make it more visually noticable and use one of the plotting tools from the ggplot package.

##geom_jitter
For any ggplot we need to assign the x and y axis within the aes().

```{r geom_jitter example}
ggplot(penguins) +
  geom_jitter(mapping = aes(x = island, y = species))
```
Now we can see a quick visual of how many of each species are in each island, (Note: I added a geom_jitter for visual clarity)

Geom_jitter functions as a scatterplot but differs from (geom_point [another scatterplot feature]) in that it adds random noise, which helps with overplotting (when data points overlap) 

##geom_point
geom_point leads to scatterplots but does not induce random noise

Lets try plotting flipper length vs. body mass
```{r geom_point example 1}
ggplot(data=penguins)+
  geom_point(mapping=aes(x=flipper_length_mm, y=body_mass_g))
```
Amazing, now lets try plotting bill length vs. depth
```{r geom_point example 2}
ggplot(data=penguins) +
  geom_point(mapping=aes(x=bill_length_mm, y=bill_depth_mm))
```

##geom_smooth
geom_smooth graphs leads to smooth line plots with the datapoints; they are helpful for showing general trends in data

There is two types (called "methods") of smoothing
  *loess smoothing is best for <1000 points
          +#geom_smooth(method="loess")
  *gam smoothing (Generalized Additive Model smoothing): best for large # of points
          +#geom_smooth(method="gam")
  R will try and compute whichever method it deems appropriate but it's good to get in the habit of noting.

```{r geom_smooth example}
ggplot(data=penguins)+
  geom_smooth(method=loess, mapping=aes(x=flipper_length_mm, y=body_mass_g))
```

##combining geom_point and geom_smooth
GGplot visualizations can be piled on.

For example geom_point and geom_smooth can also be combined in a singular visualization by using the plus + feature
```{r combination geom_point and geom_smooth}
ggplot(data=penguins)+
  geom_point(mapping=aes(x=flipper_length_mm, y=body_mass_g))+
  geom_smooth(method=loess, mapping=aes(x=flipper_length_mm, y=body_mass_g))
```

##Cleaning Up Code
  If you notice your code getting a bit complicated (especially with two types of ggplot2) you can change where you place you code of the aes (aestetics) to the section in the ggplot like below:
  
```{r cleaning up code}
ggplot(penguins, aes(x=flipper_length_mm, y=body_mass_g)) + 
  geom_point()+
  geom_smooth(method=loess)#note even if you move the aes you should still classify the method for geom_smooth
  
```

##Facet-ing Data
If you would like separate the information of the visualization out we can use the facet_wrap(~) and then which ever parameter you'd like to seperate by

```{r facet_wrap example}
ggplot(penguins, aes(x=flipper_length_mm, y=body_mass_g)) +
  geom_point()+
  geom_smooth(method=loess)+
  facet_wrap(~species)
```

#Labeling and Annotating Graphs and Other Aestetics
##Identifying Groups in Data
To identify a set of data from your code (for example type of species) we can add that information within our aes area. We can use color, size, and shape to distinguish

In the example below I used color:
```{r aes color}
ggplot(penguins, aes(x=flipper_length_mm, y=body_mass_g, color=species)) + 
  geom_point()+
  geom_smooth(method=loess)
```
  Aestetics can pile on each other, for example if you want the species to be seprate by color and shape add a comma and connect them

```{r aes color and shape}
ggplot(penguins, aes(x=flipper_length_mm, y=body_mass_g, color=species, shape=species)) + 
  geom_point()+
  geom_smooth(method=loess)
```

The alpha function can be used as well, it shows the graph and legend with different transparencies. 

Note: this can be difficult for those with low-vision so use caution. 

```{r aes alpha}
ggplot(data=penguins)+
  geom_smooth(mapping=aes(x=flipper_length_mm, y=body_mass_g, alpha=species))
```

Another type of aestetics you can use is the linetype function which can showcase lines as different types for easier visual representation

Note: especially helpful if you are only using a black/white printed form

```{r aes linetype}
ggplot(data=penguins)+
  geom_smooth(method=loess, mapping=aes(x=flipper_length_mm, y=body_mass_g, linetype=species))
```

###Labeling
To better let people know what your plots are about labels are helpful, you can add these by using the lab function.

```{r labelling}
ggplot(penguins, aes(x=flipper_length_mm, y=body_mass_g, color=species)) + 
  geom_point()+
  geom_smooth(method=loess)+
  labs(title="Palmer Penguins: Penguin Body Breakdown", subtitle="Plot of length of flipper by body mass", caption="Data Collected by Dr. Kristen Gorman")
  
```

###Annotating
Annotations are a bit more difficult as you have to assign them to specific points, all of our aesthetics can be added to this as well

Other aesthetics can include bolding, size change, and text on an angle
```{r}
ggplot(penguins, aes(x=flipper_length_mm, y=body_mass_g,  color=species)) + 
  geom_point()+
  geom_smooth(method=loess)+
  labs(title="Palmer Penguins Body Breakdown", subtitle="Plot of length of flipper by body mass", caption="Data Collected by Dr. Kristen Gorman")+
  annotate("text", x=190,y=5000, label="The Gentoos are the largest", fontface="bold",size=4.5,angle=25)
```
For each visualization you do you might need to play around with each annotation to make sure things are clear. 

If you're finding the concept of annotating difficult you may enjoy exporting your visualization and then adding in your annotations with a different program (such as Adobe Illustrator) depending on your needs

#Quick Note for Other Important R Tools (Functions, If/Then Statements, For Loop)
Withing looking at ggplot codes you might come across other bits of R base

This is meant to be a quick note and not a full tutorial:

Functions are when you have a specific item that you then put through the function to produce a value

If/Then Statements are for when a certain condition is met [if] something else happens [then]

For Loops are for when you want to have pieces of code that repeat or can build off of eachother
  Often these are used with a combination of other items
  The general set up is usually:
      for(item in list_of items){
          do something(item)
          return()
          print()
      }
  
Lets see how these tools interact by counting the number of penguin data collected by species

```{r functions, for loop, and if/then combined}
count_species <- function(x, value) { #first turn into a variable for easy usage
                                #in the function the [value] becomes [count] when it is run
  count <- 0  #starting count at 0 value as best practice
  for (penguin in 1:length(x)) {  #begin a loop by specifying what you are looking for [penguin] within a vector [1:length(x)]
    if (x[penguin] == value)  #if (then) statements have condition [x[penguin] == value)] and then the action [count <- count + 1]
      {count <- count + 1} } #part of a loop is to allow for continuous function, this set up allows the count to be updated with each pass
  count} #this is the outcome of the function, ultimately what will be shown as the final when it's run
```

This loop within a function is then stored in a vector.
We can then use the vector to find the count for each individual species
```{r}
count_species(penguins$species, "Adelie") #the set up for this would be pulling the column name [penguins$species] and then specifying what you're looking for [Adelie]
count_species(penguins$species, "Gentoo")
count_species(penguins$species, "Chinstrap")

```

Note: There are easier ways to go about counting (as seen above) but this is a small way to see how these concepts can be used within greater bits of code

```
For more information and examples of using other ggplot functions (like geom_freqpoly and geom_boxplot): https://ggplot2-book.org/