---
title: "R_basics_walkthrough_palmer_penguins"
author: "J. Mular"
date: "2022-11-22"
output: html_document
description: Walkthrough of general R principals and code
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Revisions: 
  1/20/22 cleaned up wording and renamed to distribute as a guide, added how to pull csv file from computer
  4/18/23 Added section "Quick Note for Other Important R Tools (Functions, If/Then Statements, For Loop)"
  4/22/23 Split tutorial into multiple rmarkdown files for easy navigation

#Setting Up Environment
First thing is you have to set up your R environment by making sure the packages you need are installed
To do this use the code install.packages()

For this tutorial we'll be using 'tidyverse' and palmer penguins' and other packages to be used in future walkthroughs
```{r installing packages}
install.packages("tidyverse") #for ggplots
install.packages("palmerpenguins")
install.packages("corrplot") # for correlation plot
install.packages("GGally") # for parallel coordinate plot
install.packages("e1071") # for svm

```
Note: Palmer Penguins is a package that is directly available from CRAN and you can install it the same way you would tidyverse packages

To import csv files into RStudio, use the read.csv and copy the pathway of where it is stored on your compute (note: move your file out of downloads for best practice)
```{r read.csv example}
hotel_booking <- read.csv("C:\\Users\\jmula\\OneDrive\\Desktop\\hotel_bookings.csv")
```

To use a package you must load them first which can be done with the library() function
```{r library, loading packages}
library(tidyverse)
library(palmerpenguins)
library(corrplot) 
library(GGally) 
library(e1071)
```

#Viewing the Data of Penguins to understand what we're working with
When first working with any sort of data you want to understand the specifics of it.
```{r viewing summary of penguins}
summary(penguins)
```
Note you can also use glimpse(penguins) or head() as another way to look at the data
```{r}
glimpse(penguins)
head(penguins)
```


To see column names of a data set use colnames function
```{r viewing column names}
colnames(penguins)
```

The output shoes the column names as species, island, bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g, sex, year


#Cleaning and Coloring
##Dropping Null Data
Looking through the data we can see some open spots for some of the data, to make sure what we're working with for our questions is full data let's filter out the null data using the drop_na() function

```{r filtering out null}
penguins =  drop_na(penguins)
```
##Coloring Columns
We can see there are three main species that make up the data, since we'll be using the different species for each question we're answering let's color code them to make it a bit easier to read.

```{r colorcoding species}
pCol <- c('blue', 'red','green')
names(pCol) <- c('Gentoo', 'Adelie', 'Chinstrap') #We can now plot this data and see the master list of the penguins dataset
plot(penguins, col = pCol[penguins$species], pch = 19)
```

#Example Questions
Here are some example questions of some concepts you might encounter
##Question 1: Do the penguin species live together on the same island? If so what are the counts of each?

First let's get a count of the species on island. We can do this in two ways

```{r count of species by island}
island_counts_by_species <- penguins %>%
  group_by(island) %>%
  count(species)

View(island_counts_by_species)
```
That's one way, lets do something that's a bit more readable

```{r table of count of species}
table(penguins$species, penguins$island)
```
For our purposes the second way (table) would probably suit our needs better but other data types might be better for the first way.

From this data we can easily see there is a strong relationship between what species a penguin is and what island they are on.

Findings for this question:
Gentoo lives ONLY on Biscoe island
Chinstrap lives ONLY on Dream island
and the Adelie lives on every island counted (roughly all at the same percentage per island)

##Question 2: What are the different relational features involving sex? Are these relationships statistically significant?

###How is sex relational to species?
```{r species vs. sex table}
table(penguins$sex, penguins$species)
```
Now that we have our table how can we tell if something is significant?

We can use the chisq.test!

The chi-square test of independence evaluates whether there is an association between the categories of the two variables

It uses a null and alternative hypothesis

H0: (null hypothesis) two variables are independent (i.e. "There's no effect on the population")
H1: (alternative hypothesis) two variables are not independent (i.e. "There is an effect in the population")

To determine which hypothesis is supported we use p-value

A p-value is a statistical measurement used to validate a hypothesis against observed data.
  A p-value measures the probability of obtaining the observed results, assuming that the null hypothesis is true.
  The lower the p-value, the greater the statistical significance of the observed difference

A p-value of 0.05 or lower is generally considered statistically significant

```{r chisq.test for species v. sex}
chisq.test(table(penguins$sex, penguins$species)) 
```
Since the p-value (0.976) of the text is GREATER than 0.05, is not considered statistically significant and indicates strong evidence for the null hypothesis

Reminder: Null hypothesis means they are independent of eachother.

Conclusion: Results give evidence there is no effect of sex on the species.

Note: this is just an example of using these tools, in the official data pull it is noted the research got a near equal number of each sex for the penguins.


###How is sex relational to the island the penguin lives on?
We can use the same set up of calculations as before

```{r island vs. sex table}
table(penguins$sex, penguins$island)
```

Is it statistically significant?
```{r chisq.test for island v. sex}
chisq.test(table(penguins$sex, penguins$island)) 
```

Reminder p-value is greater than 0.05 meaning not statistically significant

Now you have enough information to move onto the next walkthrough!
